Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-23 paper.Rmd
change abstract and intro

---
title: "Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (nd).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  Their model is s outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

@@

Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story. A strong theory gives more guidance on compelling choices than does a weak theory.  Accordingly, while the choice is plausible is not necessarily compelling.  Another choice might be equally plausible.  In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.    In practical terms there is a wide array of choices that are available to instantiate any particular theory.  A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics.  One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.

@@

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups.^[Andreas Beger and Michael D. Ward, 2020, “Coup forecasts for 2020”, https://www.predictiveheuristics.com/forecasts]

@@

Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other. Even if the analysis in B&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction. Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis ad hoc models. But this is hard to do in practice. The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc. They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.

@@

More importantly, trying to contrast machine learning and theory may be misguided. Even if a "theory" section is missing, predictive models are informed by previous work and arguments. If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with cleary specified models. But we do not. What we have instead are weak theories that could be interpreted to suggest any range of equations for predicting an outcome. Instead of trying to find an instantiation of such a "theoretical" model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our weak theories.

>>>>

---
title: "Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (nd), who argue that their theory-based model can predict civil war onsets better than several atheoretical alternatives or a model with country-structural factors. We find that there are several important mistakes and that their key finding is entirely conditional on the use of parametrically smoothed ROC curves when calculating accuracy, rather than the standard empirical ROC curves that dominate the literature. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  Their model is outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine. More importantly, we argue that rather than trying to contrast the roles of theory and \"atheoretical\" machine learning in predictive modeling, it would be more productive to focus on the ways in which predictive modeling and machine learning could be used to strengthen extant theories. 
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
header-includes:
  - \usepackage{floatrow}
  - \floatsetup[figure]{capposition=bottom}
  - \usepackage[format=plain,labelfont=bf,up]{caption}
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

Although their analysis in the end does not support this point, we do not disagree that theory is important in predictive modeling. Of the various ways one could contrast ad hoc machine learning with "theoretical" modeling, B&S focus essentially on theory as a mechanism for variable selection. In this sense, theory is in fact already used, informally, in nominally ad hoc predictive modeling related to conflict forecasting. The problem is that the space of extant theories for a phenomenon like civil war is so broad and non-specific in terms of an exact specification of what variables and in which functional form should constitute a prediction model, that it's both not very clear what the "correct" model should be, nor easy to demonstrate that any particular "theoretical" model's ability to predict generalizes beyond the data that was used to create it in the first place. Rather than trying to contrast theory and machine learning as competing mechanisms for creating predictive models, it would be more productive to focus on how we can use the flexibility of machine learning models and the information predictive modeling may give us about our ability to "explain" a phenomenon to strengthen and whittle down the broad space of extant theories.

@@

Indeed most empirical social science research follows this approach: 1) delineation of an argument, 2) selection of variables that represent a plausible or compelling representation of the concepts, followed by 3) utilization of these variables in a statistical model with a dependent variable that measures the main phenomena of investigation. It is worth pointing out that this is simply one set of variables that are selected to represent the concepts in the theoretical story. A strong theory gives more guidance on compelling choices than does a weak theory.  Accordingly, while the choice is plausible it is not necessarily compelling.  Another choice might be equally plausible.  In the ICEWS example, we might include variables on intragovernmental demands (from the military toward the central government), or further look a the dynamics among opposition groups.    In practical terms there is a wide array of choices that are available to instantiate any particular theory.  A second point is that putting them in a linear regression may be a particularly crude instantiation of complicated dynamics.  One general principal that is on point in the B\&S article and is more widely relevant is that dynamical phenomenon like civil wars need to have explanatory models that change rapidly over time---and by implication that are represented by data that change fairly quickly.

@@

Conversely, not all factors that are "theoretically" important or suggested by the literature are also important for prediction, and some factors that turn out to be good predictors do not originate from prior theory. For example, both @ward:etal:2010 and @hill:jones:2014 examine a variety of variables that are considered "theoretically" important for explaining civil war onset and state repression, respectively, and find that there is substantial variation in how useful they are for for actually predicting those outcomes. And it is entirely possible that a focus on building accurate predictive models could uncover factors overlooked in the literature, e.g. the strong association between infant mortality and a broad variety of conflict outcomes like political instability [@goldstone:etal:2010], irregular leadership changes [@beger:etal:2017], and coups [@beger:ward:2020].

@@

Rather than trying to frame theoretical modeling and machine learning as competing endeavors, a more useful perspective is to consider the ways in which they can mutually reinforce each other. Even if the analysis in B&S fails to support the point, we concede that predicting on the basis of a strong theoretical model is preferable to inductive prediction. Beyond accurate predictions, such models would allow us to make causal statements that are harder to make on the basis of ad hoc models. But this is hard to do in practice. The problem is that many extant theoretical models are in fact so malleable to interpretation that they essentially are ad hoc. They are useful for identifying variables that could be useful for prediction, and this is in fact already taking place.

@@

More importantly, trying to contrast machine learning and theory may be misguided. Even if a "theory" section is missing, predictive models are informed by previous work and arguments. If we had strong general theories of something like civil war onset, we could replace more inductive, ad hoc, methods with clearly specified models. But we do not. What we have instead are weak theories that could be interpreted to suggest any range of equations for predicting an outcome. Instead of trying to find an instantiation of such a "theoretical" model that can outperform a more ad hoc alternative, we could use the flexibility of machine learning methods and insights that predictive modeling brings to improve our weak theories.
